\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-03 paper.tex
Update intro and abstract

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress, however, has led to a
% Controversial, perhaps, but true?
substantial gap opening between theory and practice.
Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
based on exhaustively detailing the ``events'' that occur
in the history of a sample,
are insufficient to describe the outputs of current methods.
We argue that the precise estimates of
\emph{all} events required
by such formalisms are fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Other subtly different concepts that we must be careful to distinguish
are the ``true'' ARG, % I don't see a better word for "true" here
describing the actual history of a sample of genomes,
and a ``population'' ARG which is
the true ARG of every individual in a population.
True ARGs are the underlying, real history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Although true population ARGs unquestionably exist,
they can also never be entirely known, in part because of a fundamental
lack of mutational information.
Even if mutation rate were high enough to uniquely identify every
recent branching point,
such a high rate would saturate the genome with mutations
and obscure deeper history.

@@

% What's the problem?
A major problem currently facing the field is that
classical mathematical formalisms and terminology
cannot adequately describe these vast inferred ARGs.
Fundamentally, these formalisms assume that an
ARG is known in complete detail
and are not suited to describing partial or approximate knowledge.
As we are actively inferring ARGs at the population scale,
and such ARGs can never be known in complete detail,
there is currently a substantial gap between our theoretical frameworks
and practical application.
The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% Not sure the point is coming across that well here, but it's a
% really important point and a significant
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent. % with recomb is surely unnecessary here
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

We address this problem by providing a simple formal definition of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompassing the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% True ARGs, being perfectly resolved, can ultimately be described in
% either formalism

% This is nice to have, but probably won't work in the final version
% without section numbers. Let's deal with that when we get there I guess.
We begin in Section~\ref{sec-gARG} by providing a precise formal definition
of a gARG, illustrated by an example ARG embedded in pedigree.
We then provide a similar definition of the classical eARG
approach in Section~\ref{sec-eARG}, and consider its limitations
in the context of current datasets and research questions.
Following this, we discuss the important concept of ``ancestral material''
in Section~\ref{sec-ancestral-material}, and how it relates
to the process of converting an eARG to a gARG.
We continue
in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
between an ARG and its local trees. Contrary to the
prevailing view, we show that a suitably encoded sequence of local trees contains
precisely as much information as the corresponding ARG.
The gARG encoding opens a rich new set of details about ARGs,
including the ideas of locally unary nodes
(Section~\ref{sec-locally-unary-edges}),
the levels of detail that can be represented in an ARG
(Section~\ref{sec-ARG-simplification}),
and the degrees of precision about recombination
that can be stored and we may seek to infer (Section~\ref{sec-precision}).
These ideas have important practical considerations,
which we illustrate by examining the qualitative properties of
ARGs inferred by four recent methods for a classical benchmark
dataset in Section~\ref{sec-example-inferred-args}.
We then discuss how the gARG framework can be
efficiently implementated in Section~\ref{sec-efficiency},
and finish with an assessment of the key challenges facing
the field in the Discussion.
Finally, the literature on ARGs is large and
confusing, and we attempt to clarify some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
a description of the Big and Little ARG stochastic processes
 (Appendix~\ref{sec-big-and-little-arg}),
a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
and a discussion of ARGs at an individual vs cell
lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

>>>>

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress, however, has led to a
% Controversial, perhaps, but true?
substantial gap opening between theory and practice.
Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
based on exhaustively detailing the ``events'' that occur
in the history of a sample,
are insufficient to describe the outputs of current methods.
Moreover, we argue that the underlying assumption
that all events can be known and precisely estimated
is fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Other subtly different concepts that we must be careful to distinguish
are the ``true'' ARG, % I don't see a better word for "true" here
describing the actual history of a sample of genomes,
and a ``population'' ARG which is
the true ARG of every individual in a population.
True ARGs are the underlying real history of a sample,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Although population-scale true ARGs unquestionably exist,
they can also never be entirely known, in part because of a fundamental
lack of mutational information.
Even if mutation rate were high enough to uniquely identify every
recent branching point,
such a high rate would saturate the genome with mutations
and obscure deeper history.

@@

% What's the problem?
A major problem currently facing the field is that
classical mathematical formalisms and terminology
cannot adequately describe these vast inferred ARGs.
Fundamentally, these formalisms assume that an
ARG is known in complete detail
and are not suited to describing partial or approximate knowledge.
As we are actively inferring ARGs at the population scale,
and such ARGs can never be known in complete detail,
there is currently a substantial gap between our theoretical frameworks
and practical application.
The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
(Note that it is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.)
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

We address this problem by providing a simple formal definition of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompasses the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% True ARGs, being perfectly resolved, can ultimately be described in
% either formalism

% This is nice to have, but probably won't work in the final version
% without section numbers. Let's deal with that when we get there I guess.
We begin in Section~\ref{sec-gARG} by providing a precise formal definition
of a gARG, illustrated by an example ARG embedded in pedigree.
We then provide a similar definition of the classical eARG
approach in Section~\ref{sec-eARG}, and consider its limitations
in the context of current datasets and research questions.
Following this, we discuss the important concept of ancestral material
in Section~\ref{sec-ancestral-material}, and how it relates
to the process of converting an eARG to a gARG.
We continue
in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
between an ARG and its local trees. Contrary to the
prevailing view, we show that a suitably encoded sequence of local trees contains
precisely as much information as the corresponding ARG.
The gARG encoding opens a rich new set of details about ARGs,
including the ideas of locally unary nodes
(Section~\ref{sec-locally-unary-edges}),
the levels of detail that can be represented in an ARG
(Section~\ref{sec-ARG-simplification}),
and the degrees of precision about recombination
that can be stored and we may seek to infer (Section~\ref{sec-precision}).
These ideas have important practical considerations,
which we illustrate by examining the qualitative properties of
ARGs inferred by four recent methods for a classical benchmark
dataset in Section~\ref{sec-example-inferred-args}.
We then discuss how the gARG framework can be
efficiently implementated in Section~\ref{sec-efficiency},
and finish with an assessment of the key challenges facing
the field in the Discussion.
Finally, the literature on ARGs is large and
confusing, and we attempt to clarify some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
a description of the Big and Little ARG stochastic processes
 (Appendix~\ref{sec-big-and-little-arg}),
a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
and a discussion of ARGs at an individual vs cell
lineage level (Appendix~\ref{sec-cell-lineages-and-args}).
